Systemix Profiling Article: Combined MAG-SMAG Analysis

### 2024-03-18 R/paper.Rmd
updated to fig. 2

suppressMessages(library(ape)) #tree basics
suppressMessages(library(treeio))
suppressMessages(library(ggtree))
suppressMessages(library(ggplot2))
suppressMessages(library(ggtreeExtra)) #geom_fruit
suppressMessages(library(ggnewscale)) #scales in ggplot2
suppressMessages(library(ggstar)) #geom_star for ggplot2
suppressMessages(library(phytools)) #MRCA
suppressMessages(library(tidyverse)) #strings, dataframes processing
suppressMessages(library(magrittr)) #pipes %>%
suppressMessages(library(tidytree)) #trees as dataframes
suppressMessages(library(RColorBrewer)) #colors palette
suppressWarnings(library(colorspace))
suppressWarnings(library(patchwork))
```

@@

plot_md_top_phyla <- ggplot(mags_md_top_phy, aes(x = reorder(phy_top, freq), y = freq)) +
  geom_bar(stat = "identity", fill = "#459682") +
  geom_text(aes(label = freq), vjust = 0.5, hjust = -0.5, size = 3) +
  labs(x = "Phylum", y = "Count") +
  scale_y_continuous(limits = c(0, 10000)) +
  theme_classic() +
  coord_flip()
ggsave("plot_md_top_phyla.pdf", plot_md_top_phyla, height = 7, width = 6)

@@

plot_depth_spec <- ggplot(mags_depth_spec) +
  geom_bar(aes(x = species, fill = phylum)) +
  labs(x = "Known biomining species", y = "Count") +
  scale_fill_manual(values = "#DD7E8B") +
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = c(0.85, 0.05),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  facet_wrap(.~ecosystem)
ggsave("plot_md_biom_species.pdf", plot_depth_spec, height = 6, width = 8)

@@

plot_depth_spec_eco <- ggplot(mags_depth_spec_eco, aes(x = reorder(ecosystem, ratio),
                                                       y = ratio, fill = ecosystem)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(ratio, 2)), vjust = 0.5, hjust = -0.5, size = 3) +
  labs(x = "Ecosystem", y = "Number of kSGBs / Number of MAGs") +
  scale_y_continuous(limits = c(0, 10)) +
  scale_fill_manual(values = mags_colors_eco) +
  theme_classic() +
  theme(legend.position = "none") +
  coord_flip()
ggsave("plot_md_species_eco.pdf", plot_depth_spec_eco, height = 7, width = 6)

@@

plot_depth_sgbs_biom <- ggplot(mags_depth_sgbs_biom,
                               aes(x = reorder(phylum, tot), y = tot, fill = cond)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(name = "", labels = c("kSGB", "uSGB"),
                    values = c("#DD7E8B", "#459682")) +
  scale_y_continuous(limits = c(0, 25)) +
  labs(x = "Phylum", y = "Number of biomining species") +
  theme_classic() +
  theme(legend.position = c(0.75, 0.2), strip.placement = "outside") +
  coord_flip()
ggsave("plot_md_SGBs_phylum_biom.pdf", plot_depth_sgbs_biom, height = 5, width = 4)

@@

# Plot tree adding phylum
plot_tree_s15_phyla <- plot_tree_s15 +
  geom_tippoint(aes(color = phylum), na.rm = TRUE, size = 1, show.legend = FALSE) +
  geom_tiplab(aes(label = phylum, color = phylum),
              size = 1.5, hjust = -0.1, show.legend = FALSE) +
  scale_colour_discrete(breaks = c('c', 'd'), na.value = NA) +
  scale_colour_manual(values = tree_colors, na.translate = FALSE) +
  theme()
  #+guides(color = guide_legend(title = "Phylum", ncol = 1, order = 1))
ggsave("plot_biom_MAGs_tree_phyla.pdf", plot_tree_s15_phyla, height = 5, width = 7)

@@

# Plot tree adding QC as multiple layers
plot_tree_s15_md <- plot_tree_s15_phyla + new_scale_fill() + new_scale_color() +
  geom_fruit(data = tree_qc_df, geom = geom_bar,
             mapping = aes(y = Bin, x = Completeness, fill = Completeness),
             orientation = "y", stat = "identity", position = "auto",
             grid.params = list(linetype = 1, size = 0.1),
             axis.params = list(axis = "x", text.size = 1.8, vjust = 1, nbreak = 3),
             pwidth = 0.15, offset = 0.35) +
  scale_fill_gradient(low = "#1F4060", high = "#4CA8F6",
                      guide = guide_legend(keyheight = 0.4, keywidth = 0.7,
                                           reverse = TRUE, order = 1)) +
  new_scale_fill() + new_scale_color() +
  geom_fruit(data = tree_qc_df, geom = geom_bar,
             mapping = aes(y = Bin, x = Contamination, fill = Contamination),
             orientation = "y", stat = "identity", position = "auto",
             grid.params = list(linetype = 1, size = 0.1),
             axis.params = list(axis = "x", text.size = 1.8, vjust = 1, nbreak = 1),
             pwidth = 0.15, offset = 0.1) +
  scale_fill_gradient(low = "#6A60A8", high = "#C4C2DE",
                      guide = guide_legend(keyheight = 0.4, keywidth = 0.7,
                                           reverse = TRUE, order = 2)) +
  new_scale_fill() + new_scale_color() +
  geom_fruit(data = tree_qc_df, geom = geom_bar,
             mapping = aes(y = Bin, x = Strain_Het, fill = Strain_Het),
             orientation = "y", stat = "identity", position = "auto",
             grid.params = list(linetype = 1, size = 0.1),
             axis.params = list(axis = "x", text.size = 1.8, vjust = 1, nbreak = 3),
             pwidth = 0.15, offset = 0.1) +
  scale_fill_gradient(name = "Strain Heterogeneity",
                      low = "#C9543F", high = "#E9bAB2",
                      guide = guide_legend(keyheight = 0.4, keywidth = 0.7,
                                           reverse = TRUE, order = 3)) +
  new_scale_fill() + new_scale_color() +
  geom_fruit(data = tree_qc_df, geom = geom_bar,
             mapping = aes(y = Bin, x = Pred_Genes, fill = Pred_Genes),
             orientation = "y", stat = "identity", position = "auto",
             grid.params = list(linetype = 1, size = 0.1),
             axis.params = list(axis = "x", text.size = 1.8, vjust = 1, nbreak = 1),
             pwidth = 0.15, offset = 0.1) +
  scale_fill_gradient(name = "Predicted Genes",
                      low = "#279854", high = "#A9DEDE",
                      guide = guide_legend(keyheight = 0.4, keywidth = 0.7,
                                           reverse = TRUE, order = 4)) +
  theme(legend.title = element_text(size = 8),
        legend.text = element_text(size = 7),
        legend.key.height = unit(3.5, units = "mm"))
ggsave("plot_biom_MAGs_tree_md.pdf", plot_tree_s15_md, height = 7, width = 10)
```

## Figure 1

```{r fig1}
# Merge generated figures using patchwork
fig1_a <- {
  (plot_tree_s15_md | plot_depth_spec_eco)
  } + plot_layout(ncol = 2, widths = c(1, 0.5))
ggsave("figures/fig1_a.pdf", fig1_a, height = 7, width = 14)

fig1_b <- {
  (plot_mags_qc / plot_depth_sgbs_biom) |
    plot_depth_sgbs_both | plot_depth_kggbs_biom
  } + plot_layout(ncol = 3, widths = c(0.5, 2, 0.75))
ggsave("figures/fig1_b.pdf", fig1_b, height = 7, width = 14)

##fig1
fig1 <- (fig1_a / fig1_b) +
  plot_annotation(tag_levels = list(c("a", "e", "b", "c", "d", "", "f")))
ggsave("figures/fig1.pdf", fig1, height = 14, width = 14)
rm(fig1_a, fig1_b)

##fig2
```

## Functional Annotation: KEGG KOs from KOfamScan and Anvio Kofam

### Heatmaps

@@

### 1. Read output file from KOfamScan hits
kegg_annot <- read.csv("/mnt/beegfs/home/mrojas/DiGenomaLab/Systemix/analysis/kegg/merge_hits/kofamscan_hits.tsv", sep = "\t")

# Remove uncharacterized and duplicates proteins
kegg_annot <- kegg_annot %>%
  filter(!grepl("uncharacterized", KO.definition) == TRUE) %>%
  distinct(KO.definition, .keep_all = TRUE)

# Extract remaining KO definitions and set KOs as row names
kegg_kos <- data.frame(KO = kegg_annot$KO, Definition = kegg_annot$KO.definition)
kegg_annot <- kegg_annot %>%
  select(-c("KO.definition")) %>%
  data.frame(row.names = 1)

# Set definitions for pathways of interest
kegg_paths <- kegg_kos %>%
  #"sulf": to group sulfur, sulfate, sulfane, sulfide, sulfo_
  #"soxr": SoxR 2Fe-2S cluster
  filter(grepl("iron|sulf|soxr|copper", Definition, ignore.case = TRUE) == TRUE) %>%
  select(KO, Definition)

## FUNCTIONS ##
## Function that returns the pathway based on the definitions
fx_setKOpathways <- function(def) {
  if (grepl("copper", def, ignore.case = TRUE) == TRUE)
    return("Copper")
  if (grepl("sulf|sox", def, ignore.case = TRUE) == TRUE)
    return("Sulfur")
  if (grepl("iron", def, ignore.case = TRUE) == TRUE)
    return("Iron")
  else
    return("Other")
}
## Function to define a list of key metabolism
fx_setKOmetabolism <- function(def) {
  #oxidation, reduction
  #fixation, hydrolysis, production, utilization, transport
  #sulfane is an oxidative process, as well as dehydrogenases
  if (grepl("oxida|oxidi|hydrog|hydrox|sulfane", def, ignore.case = TRUE) == TRUE)
    return("Oxidation")
  if (grepl("reduc", def, ignore.case = TRUE) == TRUE)
    return("Reduction")
  if (grepl("utili", def, ignore.case = TRUE) == TRUE)
    return("Utilization")
  if (grepl("hydrolase", def, ignore.case = TRUE) == TRUE)
    return("Hydrolysis")
  if (grepl("transferase", def, ignore.case = TRUE) == TRUE)
    return("Transferase")
  if (grepl("transport|binding", def, ignore.case = TRUE) == TRUE)
    return("Transport")
  if (grepl("resistance", def, ignore.case = TRUE) == TRUE)
    return("Resistance")
  if (grepl("carrier|chaperone", def, ignore.case = TRUE) == TRUE)
    return("Carrier")
  if (grepl("transcriptional|regulator", def, ignore.case = TRUE) == TRUE)
    return("Regulator")
  if (grepl("lyase", def, ignore.case = TRUE) == TRUE)
    return("Lyase")
  else
    return("Other")
}
## Function that returns the ecosystem given a MAG
fx_getMAGecosystem <- function(mag) {
  if (grepl("spades", mag, ignore.case = TRUE) == TRUE) {
    return("Biomining")
  }
  else {
    mags_df <- mags_metadata_bac %>%
      dplyr::mutate(bin = paste0(mag_name, ".", mag_id)) %>%
      dplyr::filter(bin == mag) %>%
      dplyr::select(ecosystem) %>%
      as.character()
  }
}
## END FUNCTIONS ##

# Save pathways and metabolism to dataframe
kegg_ko_paths <- vector()
kegg_ko_metab <- vector()
for (i in 1:nrow(kegg_paths)) {
	kegg_ko_paths <- append(kegg_ko_paths, fx_setKOpathways(kegg_paths$Definition[i]))
	kegg_ko_metab <- append(kegg_ko_metab, fx_setKOmetabolism(kegg_paths$Definition[i]))
}
kegg_paths['Pathway'] <- kegg_ko_paths
kegg_paths['Metabolism'] <- kegg_ko_metab

# Filter matrix to select key pathways
kegg_rows <- rowSums(kegg_annot)
#summary(kegg_rows)
#table(kegg_rows)
kegg_filt <- kegg_annot[row.names(kegg_annot) %in% kegg_paths$KO, ]

# Normalize filtered rows by z-score
kegg_filt_norm <- apply(kegg_filt, 1,
                        norm <- function(x){return (x - mean(x)) / sd(x)}) %>% t()

#(plot test)
heatmap(as.matrix(kegg_filt))
heatmap(kegg_filt_norm)

@@

# Annotate rows: KO pathways
kegg_hm_rows_path <- c("Iron" = "#666666", "Sulfur" = "#DDAD3B", "Copper" = "#CA6627")
kegg_hm_rows_path <- list(Def = kegg_paths$Pathway, Pathways = kegg_hm_rows_path)

kegg_hm_rows_ann_path <- HeatmapAnnotation(Pathways = kegg_paths$Pathway,
                                           col = kegg_hm_rows_path, border = FALSE,
                                           show_annotation_name = FALSE,
                                           show_legend = TRUE, which = "row")

# Annotate rows: KO metabolism
kegg_hm_rows_metab <- setNames(brewer.pal(length(unique(kegg_paths$Metabolism)), "Paired"), unique(kegg_paths$Metabolism)) #Set3
kegg_hm_rows_metab <- sample(kegg_hm_rows_metab)
kegg_hm_rows_metab <- list(Met = kegg_paths$Metabolism, Metabolism = kegg_hm_rows_metab)

kegg_hm_rows_ann_metab <- HeatmapAnnotation(Metabolism = kegg_paths$Metabolism,
                                            col = kegg_hm_rows_metab, border = FALSE,
                                            show_annotation_name = FALSE,
                                            show_legend = TRUE, which = "row")

# Annotate cols: separate biomining MAGs from other soil SMAG
kegg_mags <- colnames(kegg_filt) %>% as.data.frame()
colnames(kegg_mags) <- c("mag")

kegg_mags_eco <- vector()
for (i in 1:nrow(kegg_mags)) {
	kegg_mags_eco <- append(kegg_mags_eco, fx_getMAGecosystem(kegg_mags$mag[i]))
}
kegg_mags['ecosystem'] <- kegg_mags_eco
#table(kegg_mags$ecosystem)

kegg_hm_cols_mags <- setNames(brewer.pal(length(unique(kegg_mags$ecosystem)), "Paired"), unique(kegg_mags$ecosystem))
kegg_hm_cols_mags <- sample(kegg_hm_cols_mags)
kegg_hm_cols_mags <- list(Eco = kegg_mags$ecosystem, Ecosystem = kegg_hm_cols_mags)

@@

# Heatmap with annotations
pdf("plot_annot_heatmap_kofam.pdf", height = 9, width = 9)
kegg_heatmap <- Heatmap(kegg_filt_norm, name = "Abundance",
                        row_title = "KEGG KOs", column_title = "Ecosystem MAGs",
                        show_row_names = FALSE, show_column_names = FALSE,
                        col = kegg_hm_colors, border = FALSE,
                        row_split = kegg_paths$Pathway,
                        column_split = kegg_mags$ecosystem,
                        right_annotation = kegg_hm_rows_ann_metab,
                        left_annotation = kegg_hm_rows_ann_path,
                        top_annotation = kegg_hm_cols_ann_mags)
draw(kegg_heatmap, merge_legend = TRUE)
while (!is.null(dev.list())) dev.off()

### 2. Read output file from Anvio Kofam
anvio_annot <- read.csv("/mnt/beegfs/home/mrojas/DiGenomaLab/Systemix/analysis/kegg/results/anvio_kofam_hits.tsv", sep = "\t")

# Save KO definitions and set row names
#anvio_kos <- data.frame(KO = anvio_annot$ko,
#                        Modules = anvio_annot$modules_with_ko,
#                        Definition = anvio_annot$ko_definition)
#anvio_annot <- anvio_annot %>% select(-c("modules_with_ko", "ko_definition"))
#anvio_annot <- data.frame(anvio_annot, row.names = 1)
#
## Set definition for selected pathways
#anvio_paths <- anvio_kos %>%
#  filter(grepl("iron|sulfur|copper", Definition, ignore.case = TRUE) == TRUE) %>%
#  select(KO, Definition)
##which(anvio_paths$KO %in% kegg_paths$KO)
#
#anvio_def_paths <- vector()
#for (i in 1:nrow(anvio_paths)) {
#	anvio_def_paths <- append(anvio_def_paths, setKODefPathway(anvio_paths$Definition[i]))
#}
#anvio_paths['Pathway'] <- anvio_def_paths
#
## Filter matrix to select key pathways
## Preprocess matrix.. (filter KOs per row)
#anvio_rows <- rowSums(anvio_annot)
#summary(anvio_rows)
#table(anvio_rows)
#anvio_filt <- anvio_annot[row.names(anvio_annot) %in% anvio_paths$KO, ]
```

@@

### 1. KOfamScan

# Note: relevant cycles (iron, sulfur, copper) have low representation in row sums,
# so during filtration we need to consider them within the range of values.
# For example, rows with low variances could be kept but they are also too many..

# Preprocess matrix
kegg_pca_mat <- kegg_annot[kegg_rows > 1 & kegg_rows < round(summary(kegg_rows)[4], 0), ]

# Merge KO definitions
kegg_pca_kos <- data.frame(KO = row.names(kegg_pca_mat))
kegg_pca_kos <- left_join(kegg_pca_kos, kegg_kos, by = c("KO"))

@@

# Normalize data to z-score
kegg_pca_mat <- apply(kegg_pca_mat, 1,
                      norm_pca <- function(x){return (x - mean(x)) / sd(x)}) %>% t()

# PCA based on SVD
set.seed(12345)
kegg_pca <- prcomp(kegg_pca_mat, center = FALSE, scale. = FALSE)

@@

# Screeplot
kegg_pca_scree_df <- data.frame(PC = 1:ncol(kegg_pca$rotation), Var = kegg_pca_var_exp)

>>>>

suppressMessages(library(ape)) #tree basics
suppressMessages(library(treeio))
suppressMessages(library(ggtree))
suppressMessages(library(ggplot2))
suppressMessages(library(ggtreeExtra)) #geom_fruit
suppressMessages(library(ggnewscale)) #scales in ggplot2
suppressMessages(library(ggstar)) #geom_star for ggplot2
suppressMessages(library(phytools)) #MRCA
suppressMessages(library(tidyverse)) #strings, dataframes processing
suppressMessages(library(magrittr)) #pipes %>%
suppressMessages(library(tidytree)) #trees as dataframes
suppressMessages(library(RColorBrewer)) #colors palette
suppressWarnings(library(colorspace))
suppressWarnings(library(patchwork)) #side-by-side plotting
```

@@

plot_depth_top_phyla <- ggplot(mags_md_top_phy, aes(x = reorder(phy_top, freq), y = freq)) +
  geom_bar(stat = "identity", fill = "#459682") +
  geom_text(aes(label = freq), vjust = 0.5, hjust = -0.5, size = 3) +
  labs(x = "Phylum", y = "Number") +
  scale_y_continuous(limits = c(0, 10000)) +
  theme_classic() +
  coord_flip()
ggsave("plot_md_top_phyla.pdf", plot_depth_top_phyla, height = 7, width = 6)

@@

plot_depth_spec <- ggplot(mags_depth_spec) +
  geom_bar(aes(x = species, fill = phylum)) +
  labs(x = "", y = "Number of known species") +
  scale_fill_manual(values = "#DD7E8B") +
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = c(0.85, 0.05),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  facet_wrap(.~ecosystem)
ggsave("plot_md_biom_species.pdf", plot_depth_spec, height = 6, width = 8)

@@

plot_depth_spec_eco <- ggplot(mags_depth_spec_eco, aes(x = reorder(ecosystem, ratio),
                                                       y = ratio, fill = ecosystem)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(ratio, 2)), vjust = 0.5, hjust = -0.5, size = 3) +
  labs(x = "Ecosystem", y = "Number of kSGBs / MAGs") +
  scale_y_continuous(limits = c(0, 10)) +
  scale_fill_manual(values = mags_colors_eco) +
  theme_classic() +
  theme(legend.position = "none") +
  coord_flip()
ggsave("plot_md_species_eco.pdf", plot_depth_spec_eco, height = 7, width = 6)

@@

plot_depth_sgbs_biom <- ggplot(mags_depth_sgbs_biom,
                               aes(x = reorder(phylum, tot), y = tot, fill = cond)) +
  geom_col(position = position_stack(reverse = TRUE)) +
  scale_fill_manual(name = "", labels = c("kSGB", "uSGB"),
                    values = c("#DD7E8B", "#459682")) +
  scale_y_continuous(limits = c(0, 25)) +
  labs(x = "Phylum", y = "Biomining species") +
  theme_classic() +
  theme(legend.position = c(0.75, 0.2), strip.placement = "outside") +
  coord_flip()
ggsave("plot_md_SGBs_phylum_biom.pdf", plot_depth_sgbs_biom, height = 5, width = 4)

@@

# Plot tree adding phylum
plot_tree_s15_phyla <- plot_tree_s15 +
  geom_tippoint(aes(color = phylum), na.rm = TRUE, size = 1) +
  geom_tiplab(aes(label = phylum, color = phylum), size = 1.75, hjust = -0.1) +
  scale_colour_discrete(breaks = c("c", "d"), na.value = NA) +
  scale_colour_manual(values = tree_colors, na.translate = FALSE, guide = "none") +
  #guides(color = guide_legend(title = "Phyla", ncol = 1, order = 1)) +
  theme()
ggsave("plot_biom_MAGs_tree_phyla.pdf", plot_tree_s15_phyla, height = 5, width = 7)

@@

# Plot tree adding QC as multiple layers
plot_tree_s15_md <- plot_tree_s15_phyla +
  new_scale_fill() + new_scale_color() +
  geom_fruit(data = tree_qc_df, geom = geom_bar,
             mapping = aes(y = Bin, x = Completeness, fill = Completeness),
             orientation = "y", stat = "identity", position = "auto", width = 0.85,
             grid.params = list(linetype = 1, size = 0.1), pwidth = 0.2, offset = 0.6,
             axis.params = list(axis = "x", text.size = 1.8, vjust = 1, nbreak = 3,
                                title = "Complet.", title.size = 1.6,
                                title.height = 0.035, title.angle = 30)) +
  scale_fill_gradient(low = "#1F4060", high = "#4CA8F6", guide = "none") +
                      #guide = guide_legend(keyheight = 0.4, keywidth = 0.7,
                      #                     reverse = TRUE, order = 1)) +
  ###Error in scale$guide == "none"
  new_scale_fill() + new_scale_color() +
  geom_fruit(data = tree_qc_df, geom = geom_bar,
             mapping = aes(y = Bin, x = Contamination, fill = Contamination),
             orientation = "y", stat = "identity", position = "auto", width = 0.85,
             grid.params = list(linetype = 1, size = 0.1), pwidth = 0.2, offset = 0.1,
             axis.params = list(axis = "x", text.size = 1.8, vjust = 1, nbreak = 1,
                                title = "Contam.", title.size = 1.6,
                                title.height = 0.035, title.angle = 30)) +
  scale_fill_gradient(low = "#6A60A8", high = "#C4C2DE", guide = "none") +
  new_scale_fill() + new_scale_color() +
  geom_fruit(data = tree_qc_df, geom = geom_bar,
             mapping = aes(y = Bin, x = Strain_Het, fill = Strain_Het),
             orientation = "y", stat = "identity", position = "auto", width = 0.85,
             grid.params = list(linetype = 1, size = 0.1), pwidth = 0.2, offset = 0.1,
             axis.params = list(axis = "x", text.size = 1.8, vjust = 1, nbreak = 3,
                                title = "Strain Het.", title.size = 1.6,
                                title.height = 0.035, title.angle = 30)) +
  scale_fill_gradient(low = "#C9543F", high = "#E9BAB2",
                      name = "Strain Heterogeneity", guide = "none") +
  new_scale_fill() + new_scale_color() +
  geom_fruit(data = tree_qc_df, geom = geom_bar,
             mapping = aes(y = Bin, x = Pred_Genes, fill = Pred_Genes),
             orientation = "y", stat = "identity", position = "auto", width = 0.85,
             grid.params = list(linetype = 1, size = 0.1), pwidth = 0.2, offset = 0.1,
             axis.params = list(axis = "x", text.size = 1.8, vjust = 1, nbreak = 1,
                                title = "Pred. genes", title.size = 1.6,
                                title.height = 0.035, title.angle = 30)) +
  scale_fill_gradient(low = "#279854", high = "#A9DEDE",
                      name = "Predicted Genes", guide = "none") +
  theme()
  #theme(legend.title = element_text(size = 8),
  #      legend.text = element_text(size = 7),
  #      legend.key.height = unit(3.5, units = "mm"),
  #      legend.key.width = unit(2, units = "mm"))
ggsave("plot_biom_MAGs_tree_md.pdf", plot_tree_s15_md, height = 7, width = 7) #x9
```

## Figure 1 & supplementary 1

```{r fig1}
# Merge plots using patchwork
fig_1a <- { plot_tree_s15_md | plot_depth_spec_eco +
  plot_layout(ncol = 2, widths = c(1, 0.5))
  } + plot_annotation(tag_levels = list(c("a", "e"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_1a.pdf", fig_1a, height = 7, width = 14)

fig_1b <- { plot_mags_qc / plot_depth_sgbs_biom |
    plot_depth_sgbs_both | plot_depth_kggbs_biom
  } + plot_layout(ncol = 3, widths = c(0.5, 2, 0.75)) +
  plot_annotation(tag_levels = list(c("b", "c", "d", "", "f"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_1b.pdf", fig_1b, height = 7, width = 14)

fig_1 <- (fig_1a / fig_1b) +
  plot_annotation(tag_levels = list(c("a", "e", "b", "c", "d", "", "f"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_1.pdf", fig_1, height = 14, width = 14)
rm(fig_1a, fig_1b)
#ggsave("fig_1.png", fig_1, dpi = 300, dev = "png", height = 14, width = 14, units = "cm")

##fig sup 1
fig_s1 <- { plot_depth_top_phyla | plot_depth_spec } +
  plot_layout(ncol = 2, widths = c(0.5, 1)) &
  theme(axis.title.x = element_text(margin = margin(t = -75, unit = "pt")))
fig_s1 <- { fig_s1 /
    { (plot_spacer() + plot_depth_usgbs + plot_spacer()) +
        plot_layout(ncol = 3, widths = c(0.5, 12, 2)) } #centered
  } + plot_annotation(tag_levels = list(c("a", "b", "c"))) &
  theme(plot.tag =  element_text(face = "bold"))
ggsave("fig_s1.pdf", fig_s1, height = 14, width = 14)
#ggsave("fig_s1.png", fig_s1, dpi = 300, dev = "png", height = 14, width = 14, units = "cm")
```

## Functional Annotation: KEGG KOs from KOfamScan

### Heatmap

@@

# Read output file from KOfamScan
kegg_annot <- read.csv("/mnt/beegfs/home/mrojas/DiGenomaLab/Systemix/analysis/kegg/merge_hits/kofamscan_hits.tsv", sep = "\t")

# Remove uncharacterized and duplicated proteins
kegg_annot <- kegg_annot %>%
  dplyr::filter(!grepl("uncharacterized", KO.definition) == TRUE) %>%
  distinct(KO.definition, .keep_all = TRUE)

# Extract KO definitions and set KOs as row names
kegg_kos <- data.frame(KO = kegg_annot$KO, Definition = kegg_annot$KO.definition)
kegg_annot <- kegg_annot %>%
  dplyr::select(-c("KO.definition")) %>%
  data.frame(row.names = 1)

# Set definitions for pathways of interest
kegg_paths <- kegg_kos %>%
  #"sulf": to group sulfur, sulfate, sulfane, sulfide, sulfo_
  #"soxr": SoxR 2Fe-2S cluster
  dplyr::filter(grepl("iron|sulf|soxr|copper", Definition, ignore.case = TRUE) == TRUE) %>%
  dplyr::select(KO, Definition)

## FUNCTIONS ##
# Function that returns the pathway based on definitions
fx_setKOpathways <- function(def) {
  if (grepl("copper", def, ignore.case = TRUE) == TRUE)
    return("Copper")
  if (grepl("sulf|sox", def, ignore.case = TRUE) == TRUE)
    return("Sulfur")
  if (grepl("iron", def, ignore.case = TRUE) == TRUE)
    return("Iron")
  else
    return("Other")
}
# Function to define types of metabolism
fx_setKOmetabolism <- function(def) {
  #oxidation, reduction
  #fixation, hydrolysis, production, utilization, transport
  #sulfane is an oxidative process, as well as dehydrogenases
  if (grepl("oxida|oxidi|hydrog|hydrox|sulfane", def, ignore.case = TRUE) == TRUE)
    return("Oxidation")
  if (grepl("reduc", def, ignore.case = TRUE) == TRUE)
    return("Reduction")
  if (grepl("utili", def, ignore.case = TRUE) == TRUE)
    return("Utilization")
  if (grepl("hydrolase", def, ignore.case = TRUE) == TRUE)
    return("Hydrolysis")
  if (grepl("transferase", def, ignore.case = TRUE) == TRUE)
    return("Transferase")
  if (grepl("transport|binding", def, ignore.case = TRUE) == TRUE)
    return("Transport")
  if (grepl("resistance", def, ignore.case = TRUE) == TRUE)
    return("Resistance")
  if (grepl("carrier|chaperone", def, ignore.case = TRUE) == TRUE)
    return("Carrier")
  if (grepl("transcriptional|regulator", def, ignore.case = TRUE) == TRUE)
    return("Regulator")
  if (grepl("lyase", def, ignore.case = TRUE) == TRUE)
    return("Lyase")
  else
    return("Other")
}
# Function that returns the ecosystem given a MAG
fx_getMAGecosystem <- function(mag) {
  if (grepl("spades", mag, ignore.case = TRUE) == TRUE) {
    return("Biomining")
  }
  else {
    mag_eco <- mags_metadata_bac %>%
      dplyr::mutate(bin = paste0(mag_name, ".", mag_id)) %>%
      dplyr::filter(bin == mag) %>%
      dplyr::select(ecosystem) %>%
      as.character()
    return(mag_eco)
  }
}
## END FUNCTIONS ##

# Populate both pathways and metabolism
kegg_ko_paths <- vector()
kegg_ko_metab <- vector()
for (i in 1:nrow(kegg_paths)) {
	kegg_ko_paths <- append(kegg_ko_paths, fx_setKOpathways(kegg_paths$Definition[i]))
	kegg_ko_metab <- append(kegg_ko_metab, fx_setKOmetabolism(kegg_paths$Definition[i]))
}
kegg_paths['Pathway'] <- kegg_ko_paths
kegg_paths['Metabolism'] <- kegg_ko_metab

# Keep only selected pathways from the annotation matrix
# (this matrix will be used later for bubble plots)
kegg_filt <- kegg_annot[row.names(kegg_annot) %in% kegg_paths$KO, ]

# Row-wise variance, then filter both matrix and pathways
kegg_filt_row_var <- apply(kegg_filt, 1, var)
#save in another variable!**
kegg_filt <- kegg_filt[kegg_filt_row_var >= summary(kegg_filt_row_var)[3], ] #median
kegg_paths_sub <- kegg_paths[kegg_paths$KO %in% row.names(kegg_filt), ]

# Normalize rows by z-score
kegg_filt_norm <- apply(kegg_filt, 1, norm <- function(x){return (x - mean(x)) / sd(x)}) %>% t()

#plot test
#heatmap(as.matrix(kegg_filt))
#heatmap(kegg_filt_norm)

@@

# Annotate rows: KO pathways
kegg_hm_rows_path <- c("Iron" = "#666666", "Sulfur" = "#DDAD3B", "Copper" = "#CA6627")
kegg_hm_rows_path <- list(Def = kegg_paths_sub$Pathway, Pathways = kegg_hm_rows_path)

kegg_hm_rows_ann_path <- HeatmapAnnotation(Pathways = kegg_paths_sub$Pathway,
                                           col = kegg_hm_rows_path, border = FALSE,
                                           show_annotation_name = FALSE,
                                           show_legend = TRUE, which = "row")

# Annotate rows: KO metabolism
kegg_hm_rows_metab <- setNames(brewer.pal(length(unique(kegg_paths_sub$Metabolism)), "Paired"), unique(kegg_paths_sub$Metabolism)) #Set3
kegg_hm_rows_metab <- sample(kegg_hm_rows_metab)
kegg_hm_rows_metab <- list(Met = kegg_paths_sub$Metabolism, Metabolism = kegg_hm_rows_metab)

kegg_hm_rows_ann_metab <- HeatmapAnnotation(Metabolism = kegg_paths_sub$Metabolism,
                                            col = kegg_hm_rows_metab, border = FALSE,
                                            show_annotation_name = FALSE,
                                            show_legend = TRUE, which = "row")

# Annotate cols: ecosystems
kegg_mags <- colnames(kegg_filt) %>% as.data.frame()
colnames(kegg_mags) <- c("mag")

kegg_mags_eco <- vector()
for (i in 1:nrow(kegg_mags)) {
	kegg_mags_eco <- append(kegg_mags_eco, fx_getMAGecosystem(kegg_mags$mag[i]))
}
kegg_mags['ecosystem'] <- kegg_mags_eco
rm(kegg_mags_eco)
#table(kegg_mags$ecosystem)

#match with mags_colors_eco**
kegg_hm_cols_mags <- setNames(brewer.pal(length(unique(kegg_mags$ecosystem)), "Paired"), unique(kegg_mags$ecosystem))
kegg_hm_cols_mags <- sample(kegg_hm_cols_mags)
kegg_hm_cols_mags <- list(Eco = kegg_mags$ecosystem, Ecosystem = kegg_hm_cols_mags)

@@

# Heatmap with annotations
pdf("plot_annot_heatmap_kofam.pdf", height = 9, width = 9)
kegg_heatmap <- Heatmap(kegg_filt_norm, name = "Abundance",
                        row_title = "KEGG KOs", column_title = "Ecosystem MAGs",
                        show_row_names = FALSE, show_column_names = FALSE,
                        col = kegg_hm_colors, border = FALSE,
                        row_split = kegg_paths_sub$Pathway,
                        column_split = kegg_mags$ecosystem,
                        right_annotation = kegg_hm_rows_ann_metab,
                        left_annotation = kegg_hm_rows_ann_path,
                        top_annotation = kegg_hm_cols_ann_mags)
draw(kegg_heatmap, merge_legend = TRUE)
while (!is.null(dev.list())) dev.off()
```

@@

# NOTE: selected cycles (Fe, Cu, S) have low representation in row sums,
# so during filtration we need to consider them within the range of values.
# E.g., rows with low variances could be kept but they are also too many.
#kegg_pca_annot_row_var <- apply(kegg_annot, 1, var)
#kegg_pca_dat <- kegg_annot[kegg_pca_annot_row_var > summary(kegg_pca_annot_row_var)[3], ]
kegg_pca_dat <- kegg_filt

# Merge KO definitions
kegg_pca_kos <- data.frame(KO = row.names(kegg_pca_dat))
kegg_pca_kos <- left_join(kegg_pca_kos, kegg_kos, by = c("KO"))

@@

# Normalize data to z-score
kegg_pca_dat <- apply(kegg_pca_dat, 1, norm_pca <- function(x){return (x - mean(x)) / sd(x)}) %>% t()

# PCA based on SVD
set.seed(122)
kegg_pca <- prcomp(kegg_pca_dat, center = FALSE, scale. = FALSE)

@@

# Screeplot
kegg_pca_scree_df <- data.frame(PC = 1:ncol(kegg_pca$rotation), Var = kegg_pca_var_exp)
kegg_pca_screeplot <- ggplot(kegg_pca_scree_df, aes(x = PC, y = Var)) +
  geom_bar(stat = "identity", fill = "#9cb8ca") +
  labs(x = "PC", y = "Variance Explained") +
  xlim(0, 50) + ylim(0, 0.06) +
  theme_minimal()
ggsave("plot_annot_pca_kofam_screeplot.pdf", kegg_pca_screeplot, height = 4, width = 5)
